A fast and memory-efficient transition state search method for complex energy landscapes

### 2022-06-06 manuscript.tex
Moved methods details back to main paper as appendices

\begin{document}
\title{A fast and memory-efficient transition state search method for complex energy landscapes}

@@

\date{\today}

@@

\subsection{Changes for undefined gradients}
A couple of alterations to the method must be made to account for situations where the gradients are unknown.
First, the calculation of $\kappa_d$ in \cref{eq:kd} must be adapted to avoid the use of gradients.
This can be done by simply removing the first term and just using the second term in the equation.
Secondly, L-BFGS can no longer be used because it requires knowledge of the gradients.
We must instead use a minimiser that does not require a differentiable optimisation function, for which we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by limiting the initial temperature and maximum random displacement.
We typically employ $T_0 = E_\mathrm{B} / 10$, and $d_\mathrm{max} = d(\bm{x}_1, \bm{x}_2) / 100$.

@@

\appendix
\section*{Methods}

\subsection{Adaptive discretisation test details}
Here we provide the details for the interpolations and mapping involved in the two examples demonstrating the feasibility of using an adaptive discretisation method.
For the cylindrical buckling example with a changing mesh, the resolution is refined each time the separation between the two states has halved, and is performed at the end of each iteration of the BITSS method.
This involves the number of triangles around the circumference of the cylinder increasing along the sequence: $40 \rightarrow 60 \rightarrow 80 \rightarrow 100$; with the number of degrees of freedom increasing by: $1760 \rightarrow 3960 \rightarrow 7040 \rightarrow 11000$.
The positions of the nodes on the new grid, $\{\bm{n}_i\}$, are determined by linear interpolation from the previous grid $\{\bm{p}_i\}$, using the positions of the unbuckled meshes, $\{\bm{n'}_i\}$ and $\{\bm{p'}_i\}$.
For each node of the new grid, $\bm{n'}_i$, the triangle that contains it is first identified, which we will denote $\{\bm{p'}_1,\bm{p'}_2,\bm{p'}_3\}$, and the barycentric coordinates of the point are computed, $\{\lambda_1,\lambda_2,\lambda_3\}$.
The new position is then given by $\bm{n}_i = \lambda_1 \bm{p}_1 + \lambda_2 \bm{p}_2 + \lambda_3 \bm{p}_3$.

\subsection{Energy and gradient expressions for the example systems}
\subsubsection{Particle cluster system}
In the particle cluster example system, the Lennard-Jones potential is used for the interaction between each pair of particles.
Therefore, the potential and its gradient for each pair of particles are
\begin{gather}
  E = 4\epsilon \left[ \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right], \\
  \frac{\partial E}{\partial \bm{x}_1} = -\frac{\partial E}{\partial \bm{x}_2} =
    \frac{24 \epsilon (\bm{x}_2 - \bm{x}_1)}{r^2} \left[ 2 \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right],
\end{gather}
where $\bm{x}_1$ and $\bm{x}_2$ are the positions of the two particles, $r$ is their separation, $\epsilon$ is the interaction strength, and $\sigma$ is the particle radius.

\subsubsection{Cylindrical buckling system}
A 2D triangular mesh is used to model the cylindrical buckling system, with the ends of the cylinder fixed in place to apply an axial compression of 0.14\%.
The energy of the system is evaluated by treating all bonds in the mesh as an elastic spring to obtain the stretching energy, and all pairs of adjacent triangles to be connected by elastic hinges, providing the bending energy.
Their expressions are given by
\begin{equation}
  E = \sum_i k^\mathrm{S}_i (r_i - r^0_i)^2 + \sum_j k^\mathrm{B}_j [1 + \cos(\theta_j - \theta^0_j)].
\end{equation}
The first term is the stretching energy, where $r_i$ is the length, $r^0_i$ is the equilibrium length, and $k^\mathrm{S}_i$ is the stretching rigidity of bond $i$.
The second term provides the bending energy, where $\theta_j$ is the dihedral angle, $\theta^0_j$ is the equilibrium angle, and $k^\mathrm{B}_j$ is the bending rigidity of hinge $j$.

\subsubsection{Striped wetting system}
This modelled on a 200x200 2D grid (and 400x400 in \cref{sec:adaptive}) using a phase-field model \cite{Panter2019b}, which has an order parameter, $\phi(\bm{r})$, representing the phase of the liquid ($\phi=-1$ for gas, $\phi=1$ for liquid).
The energy functional has four separate terms,
\begin{equation} \label{eq:phasefield}
  E[\phi] = E^\mathrm{B}[\phi] + E^\mathrm{I}[\phi] + E^\mathrm{S}[\phi] + E^\mathrm{V}[\phi].
\end{equation}
The first term, $E^\mathrm{B}$, uses a double well potential to set values of $\phi=\pm1$ in the bulk.
The second term then provides the interfacial energy between the liquid and gas by imposing an energy penalty to gradients in $\phi$.
$E^\mathrm{S}$ is the solid-liquid interaction energy, which sets the contact angles of the hydrophilic and hydrophobic regions to 60\si{\degree} and 110\si{\degree}, respectively.
Finally, $E^\mathrm{V}$ constrains the volume of the liquid drop by penalising any variation from the target volume.

@@

>>>>

\begin{document}
\title{A fast and memory-efficient transition state search method for complex energy landscapes}
\date{\today}

@@

@@

@@

@@

\appendix
\section{BITSS: Changes for undefined gradients}
A couple of alterations to the method must be made to account for situations where the gradients are unknown.
First, the calculation of $\kappa_d$ in \cref{eq:kd} must be adapted to avoid the use of gradients.
This can be done by simply removing the first term and just using the second term in the equation.
Secondly, L-BFGS can no longer be used because it requires knowledge of the gradients.
We must instead use a minimiser that does not require a differentiable optimisation function, for which we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by limiting the initial temperature and maximum random displacement.
We typically employ $T_0 = E_\mathrm{B} / 10$, and $d_\mathrm{max} = d(\bm{x}_1, \bm{x}_2) / 100$.

\section{Adaptive discretisation test details}
Here we provide the details for the interpolations and mapping involved in the two examples demonstrating the feasibility of using an adaptive discretisation method.
For the cylindrical buckling example with a changing mesh, the resolution is refined each time the separation between the two states has halved, and is performed at the end of each iteration of the BITSS method.
This involves the number of triangles around the circumference of the cylinder increasing along the sequence: $40 \rightarrow 60 \rightarrow 80 \rightarrow 100$; with the number of degrees of freedom increasing by: $\num{1760} \rightarrow \num{3960} \rightarrow \num{7040} \rightarrow \num{11000}$.
The positions of the nodes on the new grid, $\{\bm{n}_i\}$, are determined by linear interpolation from the previous grid $\{\bm{p}_i\}$, using the positions of the unbuckled meshes, $\{\bm{n'}_i\}$ and $\{\bm{p'}_i\}$.
For each node of the new grid, $\bm{n'}_i$, the triangle that contains it is first identified, which we will denote $\{\bm{p'}_1,\bm{p'}_2,\bm{p'}_3\}$, and the barycentric coordinates of the point are computed, $\{\lambda_1,\lambda_2,\lambda_3\}$.
The new position is then given by $\bm{n}_i = \lambda_1 \bm{p}_1 + \lambda_2 \bm{p}_2 + \lambda_3 \bm{p}_3$.

\section{Energy and gradient expressions for the example systems}
\subsection{Particle cluster system}
In the particle cluster example system, the Lennard-Jones potential is used for the interaction between each pair of particles.
Therefore, the potential and its gradient for each pair of particles are
\begin{gather}
  E = 4\epsilon \left[ \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right], \\
  \frac{\partial E}{\partial \bm{x}_1} = -\frac{\partial E}{\partial \bm{x}_2} =
    \frac{24 \epsilon (\bm{x}_2 - \bm{x}_1)}{r^2} \left[ 2 \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right],
\end{gather}
where $\bm{x}_1$ and $\bm{x}_2$ are the positions of the two particles, $r$ is their separation, $\epsilon$ is the interaction strength, and $\sigma$ is the particle radius.

\subsection{Cylindrical buckling system}
A 2D triangular mesh is used to model the cylindrical buckling system, with the ends of the cylinder fixed in place to apply an axial compression of 0.14\%.
The energy of the system is evaluated by treating all bonds in the mesh as an elastic spring to obtain the stretching energy, and all pairs of adjacent triangles to be connected by elastic hinges, providing the bending energy.
Their expressions are given by
\begin{equation}
  E = \sum_i k^\mathrm{S}_i (r_i - r^0_i)^2 + \sum_j k^\mathrm{B}_j [1 + \cos(\theta_j - \theta^0_j)].
\end{equation}
The first term is the stretching energy, where $r_i$ is the length, $r^0_i$ is the equilibrium length, and $k^\mathrm{S}_i$ is the stretching rigidity of bond $i$.
The second term provides the bending energy, where $\theta_j$ is the dihedral angle, $\theta^0_j$ is the equilibrium angle, and $k^\mathrm{B}_j$ is the bending rigidity of hinge $j$.

\subsection{Striped wetting system}
This is modelled on a 200x200 2D grid (and 400x400 in \cref{sec:adaptive}) using a phase-field model \cite{Panter2019b}, which has an order parameter, $\phi(\bm{r})$, representing the phase of the liquid ($\phi=-1$ for gas, $\phi=1$ for liquid).
The energy functional has four separate terms,
\begin{equation} \label{eq:phasefield}
  E[\phi] = E^\mathrm{B}[\phi] + E^\mathrm{I}[\phi] + E^\mathrm{S}[\phi] + E^\mathrm{V}[\phi].
\end{equation}
The first term, $E^\mathrm{B}$, uses a double well potential to set values of $\phi=\pm1$ in the bulk.
The second term then provides the interfacial energy between the liquid and gas by imposing an energy penalty to gradients in $\phi$.
$E^\mathrm{S}$ is the solid-liquid interaction energy, which sets the contact angles of the hydrophilic and hydrophobic regions to 60\si{\degree} and 110\si{\degree}, respectively.
Finally, $E^\mathrm{V}$ constrains the volume of the liquid drop by penalising any variation from the target volume.
